Clusterization Properties of D-optimal Designs in Infinite Dimensional Bayesian Linear Inverse Problems

### 2023-04-13 latex/vanishing.tex
Need to write intro. Tell how everybody thinks model error is the solution

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% SECTION Analysis of Optimal Designs --- Vanishing Model Error
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{D-Optimal Designs with Vanishing Model Error}\label{section:vanishing}
We now turn to analyze D-optimal designs when $\modcov = 0$. Our goal
is to prove Theorem \ref{thm:char} --- a characterization of D-optimal
designs.

The necessary first-order condition for D-optimality summarized in
Theorem \ref{thm:constrained}, when $\Sigma= \sigma^2I$ (i.e. when
$\modcov = 0$) becomes
\begin{equation}\label{eq:eigenproblem}
  \sigma^{-2}\fwd \postcov \fwd^* \obs^* = \obs^* \Xi,
\end{equation}
with $\Xi$ diagonal. We think of \eqref{eq:eigenproblem} as an
eigenvalue problem for the self-adjoint operator $\sigma^{-2}\fwd
\postcov \fwd^*$, where rows of $\obs$, namely $\meas_j,j=1,\dots, m$,
are eigenvectors. However, $\postcov$ depends on $\obs$ and thus, we
refer to \eqref{eq:eigenproblem} as a \emph{nonlinear} eigenvalue
problem. The following Proposition and Lemma are proved in the
appendix:

@@

\begin{proposition}[Eigenvectors of a D-optimal design]\label{prop:same ev}
  Let $\obs$ satisfy the nonlinear eigenvalue problem
  \eqref{eq:eigenproblem}. Then $\obs^*\obs$ has the same eigenvectors
  as $\fwd \prcov \fwd^*$.
\end{proposition}
\begin{proof}
  \begin{align}\label{eq:mod conditions}
    \begin{split}
      \obs^* \Xi &= \sigma^{-2}\fwd \postcov \fwd^* \obs^*  \text{ (by \eqref{eq:eigenproblem})}\\
      %
      %
      %
      &= \sigma^{-2} \fwd( \prcov^{-1} + \sigma^{-2}  \fwd^* \obs^* \obs \fwd )^{-1} \fwd^* \obs^*  \text{ (by \eqref{eq:postcov})} \\
      %
      %
      %
      &= \sigma^{-2} \left ( (\fwd\prcov\fwd^*)^{-1} + \sigma^{-2}  \obs^* \obs \right )^{-1} \obs^* \text{ (by Proposition \ref{prop:twice woodbury})}.
    \end{split}
  \end{align}

@@

We are almost ready to prove Theorem \ref{thm:char} that characterizes
D-optimal designs. We just need one final lemma, whose proof is
delegated to the appendix:
\begin{restatable*}[Unit norm decomposition]{lemma}{free}\label{lemma:free}
  Let $M \in \R^{k \times k}$ symmetric positive definite with $\ttr M
  = m$, $m \geq k$. We can find $\func_j \in \R^k,j=1,\dots,m$
  with $\|\func_j\|=1$ and $A = (\func_1,\dots,\func_m)$ such that
  $AA^t = M$.
\end{restatable*}

@@

    \end{axis}
  \end{tikzpicture}
  \caption{A comparison of the eigenvalues of the pushforward
    posterior precision $(\fwd\prcov\fwd^*)^{-1} +
    \sigma^{-2}\obs^*\obs$ for a D-optimal design (left) and a
    sub-optimal design (right). Both designs are allowed three
    observations. We assume $\sigma^2=1$ and thus, the blue area has
    accumulated height of $3$ in both plots. The D-optimal design
    (left) increases precision (equivalently, reduces uncertainty)
    where it is lowest. The sub-optimal design (right) does not.
    %% In this example, an observation is taken partly in the
    %% direction of the third eigenvector. The precision for this
    %% eigenvector is bigger even than the posterior precision of the
    %% D-optimal design on the left.
  }
  \label{fig:optimal vs not}
\end{figure}

Here is a way to understand D-optimal designs with vanishing model
error: Imagine each eigenvector of $\fwd \prcov \fwd^*$ corresponds to
a graduated lab cylinder (Figure \ref{fig:cylinder}). Each cylinder
$i$ is filled, a-priori, with green liquid of $\lambda_i^{-1}$ volume
units. This liquid represnets the prior precision, and the volume of
the green liquid is increasing in $i$. A D-optimal design with $m$
observations ignores cylinders indexed $m+1$ and above. It has blue
liquid of volume $\sigma^{-2}m$ units at its disposal. It distributes
said liquid by repeatedly adding a drop to whatever cylinder currently
has the lowest level of liquid in it (lowest precision). The result of
such a procedure is illustrated in Figure \ref{fig:optimal vs not}.

@@

\subsection{Clusterization poses no obstruction to D-optimality}\label{subsec:clusterization}
We saw that the necessary conditions for a design $\obs$ with $m$
observations to be D-optimal are (1), (2) and (4) of Theorem
\ref{thm:char}. As long as these requirements are satisfied, $\obs$ is
D-optimal. These requirements, however, still leave some freedom in
choosing $\obs$. This freedom may give rise to clusterization, as
illustrated in Figure \ref{fig:clusterization} and explained below.

Consider an inverse problem with $\lambda_1^{-1} = 0.1, \lambda_2^{-1}
= 0.2$ and $\sigma^2=1$. In \eqref{eq:two designs}, we suggest two
D-optimal designs $\obs_1, \obs_2$ such that $\obs_1^*\obs_1 =
\obs_2^*\obs_2$. The former exhibits clusterization ($\meas_3 =
\meas_5 = \ev_1$), while the latter does not. Both of these designs
are D-optimal, hence clusterization poses no obstruction to
D-optimality.

@@

A computer implementation of Lemma \ref{lemma:free} (see code in
Supplementary material) generates $\obs_1$ as a solution to the
D-optimal design problem mentioned in the previous
paragraph. Modifying the numerical values of the prior does not break
the clusterization. Thus, the preference to clusterization of Lemma
\ref{lemma:free} is robust. The vast majority of randomized numerical
simulations (see code in Supplementary material) also give rise to
D-optimal designs that exhibit clustrization. We conclude that
clusterization poses no obstruction to D-optimality in the relaxed
model presented in this study. Moreover, sensor clusterization is
generic, as far as one accepts the model presented in this study.

>>>>

\section{D-Optimal Designs when $\modcov = 0$}\label{section:vanishing}
We now turn to characterize D-optimal designs when $\modcov = 0$. Our
goal is to prove Theorem \ref{thm:char}.

The necessary first-order condition for D-optimality of Theorem
\ref{thm:constrained} for $\modcov = 0$ become:

\begin{equation}\label{eq:eigenproblem}
  \sigma^{-2}\fwd \postcov \fwd^* \obs^* = \obs^* \Xi,
\end{equation}
with $\Xi$ diagonal. Equation \eqref{eq:eigenproblem} looks like an
eigenvalue problem for the self-adjoint operator $\sigma^{-2}\fwd
\postcov \fwd^*$, where rows of $\obs$, namely $\meas_j,j=1,\dots, m$,
are eigenvectors. However, $\postcov$ depends on $\obs$ and thus, we
refer to \eqref{eq:eigenproblem} as a \emph{nonlinear} eigenvalue
problem. The following Proposition and Lemma are proved in the
appendix:

@@

\begin{proposition}[Eigenvectors of a D-optimal design]\label{prop:same ev}
  Let $\obs$ satisfy the nonlinear eigenvalue problem
  \eqref{eq:eigenproblem}. Then $\obs^*\obs$ and $\fwd \prcov \fwd^*$
  are simultaneously diagonalizable.
\end{proposition}
\begin{proof}
  \begin{align}\label{eq:mod conditions}
    \begin{split}
      \obs^* \Xi &= \sigma^{-2}\fwd \postcov \fwd^* \obs^*  \text{ (by \eqref{eq:eigenproblem})}\\
      %
      %
      %
      &= \sigma^{-2} \fwd( \prcov^{-1} + \sigma^{-2}  \fwd^* \obs^* \obs \fwd )^{-1} \fwd^* \obs^*  \text{ (by \eqref{eq:postcov})} \\
      %
      %
      %
      &= \sigma^{-2} \left ( (\fwd\prcov\fwd^*)^{-1} + \sigma^{-2}  \obs^* \obs \right )^{-1} \obs^* \text{ (by Proposition \ref{prop:twice woodbury})}.
    \end{split}
  \end{align}

@@

Before we prove Theorem \ref{thm:char} we record Lemma
\ref{lemma:free}, whose proof is delegated to the appendix:
\begin{restatable*}[Unit norm decomposition]{lemma}{free}\label{lemma:free}
  Let $M \in \R^{k \times k}$ symmetric positive definite with $\ttr M
  = m$, $m \geq k$. We can find $\func_j \in \R^k,j=1,\dots,m$
  with $\|\func_j\|=1$ and $A = (\func_1,\dots,\func_m)$ such that
  $AA^t = M$.
\end{restatable*}

@@

    \end{axis}
  \end{tikzpicture}
  \caption{A comparison of the eigenvalues of the pushforward
    posterior precision $(\fwd\prcov\fwd^*)^{-1} +
    \sigma^{-2}\obs^*\obs$ for a D-optimal design (left) and a
    sub-optimal design (right). Both designs are allowed $m=3$
    observations. We assume $\sigma^2=1$ and thus, the blue area has
    accumulated height of $\sigma^{-2}m = 3$ in both plots. The
    D-optimal design (left) increases precision (equivalently, reduces
    uncertainty) where it is lowest. The sub-optimal design (right)
    does not.}
  \label{fig:optimal vs not}
\end{figure}

Theorem \ref{thm:char} gives us a way to understand D-optimal designs
when $\modcov = 0$: Imagine each eigenvector of $\fwd \prcov \fwd^*$
corresponds to a graduated lab cylinder (Figure
\ref{fig:cylinder}). Each cylinder $i$ is filled, a-priori, with green
liquid of $\lambda_i^{-1}$ volume units. We are allowed $m$
observations, so we have blue liquid of volume $\sigma^{-2}m$ units at
our disposal. When we seek a D-optimal design, we distribute the blue
liquid by repeatedly adding a drop to whatever cylinder currently has
the lowest level of liquid in it, as long as its index $i \leq m$. The
result of such a procedure is illustrated in Figure \ref{fig:optimal
  vs not}.

@@

\section{Clusterization poses no obstruction to D-optimality}\label{section:clusterization}
The necessary conditions for D-optimality are (1), (2) and (4) of
Theorem \ref{thm:char}. As long as these conditions are satisfied,
$\obs$ is an optimizer of $\tar$. These conditions, however, still
leave some freedom in choosing $\obs$. This freedom may give rise to
clusterization, as illustrated in Figure \ref{fig:clusterization} and
demonstrated below.

Consider an inverse problem for which $\lambda_1^{-1} = 0.1,
\lambda_2^{-1} = 0.2,\sigma^2=1$ and $\{\ev_i\}_{i=1}^\infty$ are
eigenvectors of $\fwd\prcov\fwd^*$. Two D-optimal designs $\obs_1,
\obs_2$ presented in \eqref{eq:two designs}. It is easy to verify that
$\obs_1^*\obs_1 = \obs_2^*\obs_2$. The former exhibits clusterization
($\meas_3 = \meas_5 = \ev_1$), while the latter does not. Since both
of these designs are D-optimal, we conclude that clusterization poses
no obstruction to D-optimality.

@@

A computer implementation of Lemma \ref{lemma:free} (code in
Supplementary material) generates $\obs_1$ as a solution to the
D-optimal design problem mentioned in the previous
paragraph. Modifying the numerical values of the prior does not break
the clusterization. Thus, the preference to clusterization, at least
in Lemma \ref{lemma:free} is robust. The vast majority of randomized
numerical simulations (see code in Supplementary material) also give
rise to D-optimal designs that exhibit clustrization. We conclude that
clusterization poses no obstruction to D-optimality in the model
presented in this study. Moreover, sensor clusterization is generic,
as far as the model presented in this study is concerned.
